Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-04-11 mpaper/paper.tex
Editing (mostly introduction and background)

\begin{abstract}
  The inverse reinforcement learning (IRL) problem asks us to find a reward
  function of a Markov decision process that explains observed behaviour. Many
  approaches are only able to construct reward functions as linear combinations
  of state features. Out of those that can handle nonlinearity, none can provide
  a full posterior distribution of rewards. Providing variance estimates for
  rewards would allow one to judge how well the model has learned its policy and
  discover any weak spots the model may have. We show how to perform variational
  inference (VI) on a Gaussian process-based IRL model in order to approximate
  the posterior distribution of rewards. We prove the correctness of the
  approach and demonstrate the model's behaviour in practice. Being able to
  provide full posterior probability distributions in IRL unlocks many new
  research frontiers ranging from integrating recent developments in VI to make
  the models more efficient and flexible, to developing complex reinforcement
  learning agents that can explicitly search for opportunities to fix their
  weaknesses.
\end{abstract}

@@

This learning problem
\cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} as well as
many others have benefited from an approach called \emph{inverse reinforcement
  learning} (IRL) (also known as inverse optimal control). IRL proposes a way to
learn behaviour from \emph{demonstrations} that typically come from human
actions. More formally, the IRL problem asks us to find a reward function for a
Markov decision process (MDP), where demonstrations are encoded as sets of
state-action pairs.
% TODO (must): along with inverse optimal control, mention imitation learning
% and apprenticeship learning

@@

% TODO: need to rewrite the first sentence
However, most IRL models in the literature make a convenient yet unjustified
assumption that the reward function can be expressed as a linear combination of
features
\cite{DBLP:conf/icml/PieterN04,DBLP:conf/icml/NgR00,ziebart2008maximum}.
This assumption severely restricts what reward structures can be modelled (i.e.,
the properties of the function that represents rewards across states). Out of
the non-linear models proposed to date, none can answer the questions posed in
the first paragraph. Quite often, the models assume that rewards have no
variance \cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}.

In this paper, we show how that assumption can be lifted by switching from
maximum likelihood estimation to \emph{variational inference} (VI), i.e., we
approximate the posterior distribution of the model by optimising the parameters
of a simpler distribution to make it similar to the posterior. This approach can
prove useful in four major ways:
\begin{enumerate}
\item By working with full distributions instead of point estimates, we can
  expect more precise reward predictions.
\item Variance estimates can be used to guide what data should be collected
  next, i.e., if the rewards of some states have abnormally high variance, we
  might want to expose the model to more data visiting those and surrounding
  states.
\item Variances estimates can also be used to judge whether we can trust the
  predictions of the model or, perhaps, the model could benefit from some
  adjustments or more data.
\item By adopting a more Bayesian approach, we automatically incorporate Occam's
  razor into the model that guards against overfitting
  \cite{DBLP:conf/uai/JinDAS17}.
\end{enumerate}

Our main contribution is a proof in Section~\ref{sec:proof} that shows
how VI can be applied to the maximum-entropy IRL model with Gaussian processes
(GPs) proposed by Levine et al. \cite{DBLP:conf/nips/LevinePK11}. We describe
how we adapt the model and set up the VI problem in Section~\ref{sec:model}.
Finally, in Section~\ref{sec:evaluation} we examine the convergence properties
of our model and its ability to deduce optimal policies in practice.
% TODO: expand on this more

@@

\section{The Problem} \label{sec:problem}

@@

\begin{definition}[IRL]
  Given an MDP without rewards $\mathcal{M} \setminus \{ \mathbf{r} \}$, an
  $|\mathcal{S}| \times d$ feature matrix $\mathbf{X}$ (where $d$ is the number
  of features), and a set of demonstrations $\mathcal{D} = \{\zeta_i
  \}_{i=1}^N$, where each demonstration $\zeta_i = \{ (s_{i,t}, a_{i,t})
  \}_{t=1}^T$ is a multiset of state-action pairs that represents optimal (or
  near-optimal) actions, find the reward function that maximises the probability
  of observing the demonstrations, i.e.,
  \[
    \argmax_{\mathbf{r}} p(\mathcal{D} \mid \mathbf{r}).
  \]
\end{definition}

\begin{example}
  Figure \ref{fig:example} shows a possible setup for IRL. In this case,
  \begin{align*}
    \mathcal{S} &= \{ s_1, s_2, s_3, s_4 \}, \\
    \mathcal{A} &= \{ l, r, u, d \}, \\
    \mathcal{T}(s, a, s') &= \begin{cases}
      1 & \text{if } (s, a, s') = (s_1, r, s_2) \\
      1 & \text{if } (s, a, s') = (s_1, d, s_4) \\
      1 & \text{if } (s, a, s') = (s_2, l, s_1) \\
      1 & \text{if } (s, a, s') = (s_2, d, s_3) \\
      1 & \text{if } (s, a, s') = (s_3, l, s_4) \\
      1 & \text{if } (s, a, s') = (s_3, u, s_2) \\
      1 & \text{if } (s, a, s') = (s_4, r, s_3) \\
      1 & \text{if } (s, a, s') = (s_4, u, s_1) \\
      0 & \text{otherwise,}
    \end{cases} \\
      \mathbf{X} &= \begin{blockarray}{ccc}
        & f_1 & f_2 \\
        \begin{block}{c[cc]}
          s_1 & 1 & 2 \\
          s_2 & 2 & 2 \\
          s_3 & 2 & 1 \\
          s_4 & 1 & 1 \\
        \end{block}
      \end{blockarray},
  \end{align*}
  and $\gamma \in [0, 1)$ can be assigned any value. The distribution of example
  state-action pairs into separate demonstrations is irrelevant to our model
  (and possible all other models), so we can keep each pair in its own set as
  follows:
  \[
    \mathcal{D} = \{ \{(s_1, d)\}, \{(s_2, l)\}, \{(s_4, u)\}, \{(s_4, u)\} \}.
  \]

@@

A common way to scale GPs to larger data sets is by selecting $m$ points in the
feature space--called \emph{inducing points}---and focus most of the training
effort on them \cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be
the $m \times d$ matrix of features at inducing points, and let $\mathbf{u}$ be
the rewards at those states. Then the full joint probability distribution can be
factorised as
\begin{equation} \label{eq:full}
  \pfull = p(\mathbf{u}) \times p(\mathbf{r} \mid \mathbf{u}) \times
  p(\mathcal{D} \mid \mathbf{r}),
\end{equation}
where
\begin{align*}
  p(\mathbf{u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \\
                &= \frac{1}{(2\pi)^{m/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} \right) \\
                &= \exp\left(-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right)
\end{align*}
is the GP prior \cite{DBLP:books/lib/RasmussenW06}. The GP posterior is then a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11} defined as
\begin{equation} \label{eq:r}
  p(\mathbf{r} \mid \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru).
\end{equation}
The matrices such as
$\Kru$ are called \emph{covariance matrices} (also \emph{kernel matrices} and
\emph{Gram matrices}) and are defined as
$[\Kru]_{i,j} = k(\mathbf{x}_{\mathbf{r},i},
\mathbf{x}_{\mathbf{u},j})$, where $\mathbf{x}_{\mathbf{r},i}$ and
$\mathbf{x}_{\mathbf{u},j}$ denote feature vectors for the $i$th state in
$\mathcal{S}$ and the $j$th inducing point, respectively
\cite{DBLP:conf/uai/JinDAS17}.

Our goal is then to use VI in order to approximate
$p(\mathbf{u},~\mathbf{r}~\mid~\mathcal{D})$. Let $\approximation$ denote our
approximation. VI aims to optimise this approximation by minimising the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation. KL divergence (asymmetrically) measures the
difference between two probability distributions and is defined as
\cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}_{\approximation}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}_{\approximation}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}_{\approximation}[\log p(\mathcal{D})].
\end{align*}
The last term is both hard to compute and constant with respect to
$\approximation$ \cite{blei2017variational}, so we can remove it from our
optimisation objective. The negation of what remains is known as the
\emph{evidence lower bound} (ELBO) and is defined as
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}_{\approximation}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}
Thus, instead of minimising KL divergence, we focus on maximising $\mathcal{L}$
by optimising the values of parameters to be defined in Section~\ref{sec:model}.
Note that hereafter we drop the subscript notation, as all expected values will
be with respect to $\approximation$.

\section{Background}

We introduce a few definitions and results from linear algebra, numerical
analysis, and measure theory that will be used later in the paper. Namely, we
will use several different vector and matrix norms, consider how an inverse of a
matrix changes with a small perturbation, and introduce Lebesgue's dominated
convergence theorem which plays a major role in our theoretical results in
Section~\ref{sec:proof}.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[latent] (u) {$\mathbf{u}$};
    \node[latent, right=of u] (r) {$\mathbf{r}$};
    \node[obs, right=of r] (D) {$\mathcal{D}$};
    \node[const, below=of r] (lambda) {$\bm\lambda$};
    \node[const, below=of u] (mu) {$\bm\mu$};
    \node[const, left=of mu] (B) {$\mathbf{B}$};
    \edge {mu, B} {u};
    \edge {lambda, u} {r};
    \edge {r} {D};
  \end{tikzpicture}
  \caption{Our VI problem expressed as a (simplified)
    Bayesian network. The only observed variable (representing the
    demonstrations) is in a gray circle, modelled latent variables are in white
    circles, and the variational parameters are at the bottom.}
  \label{fig:graphical_model}
\end{figure}

@@

Firstly, let us return to Equation~\eqref{eq:elbo} and write
\[
  \mathcal{L} = \mathbb{E}[\log\pfull] - \mathbb{E}[\log\approximation].
\]
By substituting in Equations~\eqref{eq:full} and \eqref{eq:approximation}, we get
\begin{align*}
  \mathcal{L} &= \mathbb{E}[\log p(\mathbf{u}) + \log p(\mathbf{r} \mid \mathbf{u}) + \log p(\mathcal{D} \mid \mathbf{r})] \\
              &- \mathbb{E}[\log q(\mathbf{u}) + \log q(\mathbf{r} \mid \mathbf{u})].
\end{align*}
Since $q(\mathbf{r} \mid \mathbf{u}) = p(\mathbf{r} \mid \mathbf{u})$, they
cancel each other out. Also notice that
\begin{align*}
  \mathbb{E}[\log p(\mathbf{u}) - \log q(\mathbf{u})] &= -\DKL{q(\mathbf{u})}{p(\mathbf{u})} \\
                                                      &= -\frac{1}{2} (\tr (\Kuu^{-1}\bm\Sigma) + \bm\mu^\intercal\Kuu^{-1}\bm\mu - m \\
                                                      &+ \log |\Kuu| - \log |\bm\Sigma|),
\end{align*}
by the definition of KL divergence between two multivariate Gaussians
\cite{kl}. Hence,
\begin{align*}
  \mathcal{L} &= \mathbb{E}\left[ \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right).
\end{align*}
Using the expressions for $Q_{\mathbf{r}}$ we get
\begin{align*}
  \mathcal{L} &= \mathbb{E}\left[\sum_{i=1}^N \sum_{t=1}^T r(s_{i,t}) - \V(s_{i,t}) + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t}, s')\V(s') \right] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right).
\end{align*}
We can simplify $\sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t})$ by defining a new vector
$\mathbf{t} = (t_1, \dots, t_{|\mathcal{S}|})^\intercal$, where $t_i$ is the
number of times the state associated with reward $r_i$ has been visited
across all demonstrations. Then,
\begin{align*}
  \mathbb{E} \left[ \sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t}) \right] &= \mathbb{E}[\mathbf{t}^\intercal\mathbf{r}] = \mathbf{t}^\intercal\mathbb{E}[\mathbf{r}] \\
                                                               &= \mathbf{t}^\intercal\mathbb{E}\left[\Kru^\intercal\Kuu^{-1}\mathbf{u}\right] = \mathbf{t}^\intercal\Kru^\intercal\Kuu^{-1}\bm\mu.
\end{align*}
This allows us to simplify $\mathcal{L}$ to
\begin{align*}
  \mathcal{L} &= \mathbf{t}^\intercal\Kru^\intercal\Kuu^{-1}\bm\mu - \mathbb{E}[v] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right),
\end{align*}
where
\[
  v = \sum_{i=1}^N \sum_{t=1}^T \V(s_{i,t}) - \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s_{i,t}, a_{i,t}, s')\V(s').
\]

\section{Theoretical Justification} \label{sec:proof}
% TODO: Possible improvements:
% 1. Use limits more often.
% 2. Avoid saying that epsilon is arbitrary and using phrases like 'sufficiently small'.

The typical way to optimise a quantity (the ELBO, in this case) involves
computing its gradient. Unfortunately, the term $\mathbb{E}[v]$ in $\mathcal{L}$
complicates the situation. The goal of this section is to show how
Theorem~\ref{thm:lebesgue} can be applied to our model in order to derive the
gradient anyway\footnote{This technique is inspired by black box VI
  \cite{DBLP:conf/aistats/RanganathGB14}, but takes a more detailed look at the
  problem and requires significantly more work to prove correctness.}. After
showing that the theorem applies to our situation, we can estimate
$\nabla\mathbb{E}[v]$ with
\begin{align*}
  \nabla\mathbb{E}[v] &= \nabla \iint v q(\mathbf{r} \mid \mathbf{u}) q(\mathbf{u})\dx \\
                      &= \iint \nabla[v q(\mathbf{r} \mid \mathbf{u}) q(\mathbf{u})]\dx \\
                      &= \iint \frac{\nabla[v q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})]}{q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})} q(\mathbf{r} \mid \mathbf{u}) q(\mathbf{u})\dx \\
                      &\approx \frac{1}{S} \sum_{s=1}^S \frac{\nabla[v q(\mathbf{r}_s \mid \mathbf{u}_s)q(\mathbf{u}_s)]}{q(\mathbf{r}_s \mid \mathbf{u}_s)q(\mathbf{u}_s)},
\end{align*}
which can be computed by drawing $S$ samples $\{(\mathbf{u}_s,
\mathbf{r}_s)\}_{s=1}^S$ from $\approximation$.

Our main goal is Theorem~\ref{thm:main}, which allows us to move differentiation
inside the integral. In order to prove it, we use a number of intermediate
results. We start by stating a few derivatives of PDFs and covariance matrices,
and bound their values with some easy-to-deal-with polynomials. We then provide
a sketch proof of the measurability of MDP value functions as well as new upper
and lower bounds on their values. After another quick lemma, we tackle the main
proof of this paper. See Figure~\ref{fig:theorems} for an overview of how these
results fit together.

@@

>>>>

\begin{abstract}
  The inverse reinforcement learning (IRL) problem asks us to find a reward
  function of a Markov decision process that explains observed behaviour. Many
  approaches are only able to express reward functions as linear combinations
  of state features. Out of those that can handle nonlinearity, none can provide
  a full posterior distribution of rewards. Providing variance estimates for
  rewards would allow one to judge how well the model has learned its policy and
  discover any weak spots the model may have. We show how to perform variational
  inference (VI) on a Gaussian process-based IRL model in order to approximate
  the posterior distribution of rewards. We prove the correctness of the
  approach and demonstrate the model's behaviour in practice. Being able to
  provide full posterior probability distributions in IRL unlocks many new
  research frontiers ranging from integrating recent developments in VI to make
  the models more efficient and flexible, to developing complex reinforcement
  learning agents that can explicitly search for opportunities to fix their
  weaknesses.
\end{abstract}

@@

This learning problem
\cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} as well as
many others have benefited from an approach called \emph{inverse reinforcement
  learning} (IRL), also known as apprenticeship learning and inverse optimal
control. IRL proposes a way to learn behaviour from demonstrations that
typically come from human actions. More formally, the IRL problem asks us to
find a reward function for a Markov decision process (MDP), where demonstrations
are encoded as sets of state-action pairs.

@@

Most IRL models in the literature make a convenient yet unjustified
assumption that the reward function can be expressed as a linear combination of
features
\cite{DBLP:conf/icml/PieterN04,DBLP:conf/icml/NgR00,ziebart2008maximum}.
This assumption severely restricts what reward structures can be modelled (i.e.,
the properties of the function that represents rewards across states). Out of
the non-linear models proposed to date, none can answer the questions posed in
the first paragraph. More specifically, without knowing the ground truth for the
learning problem, we have no reliable way to know whether the learned behaviour
is optimal and which parts of the model could benefit from more data. Quite
often, the models assume that rewards have no variance
\cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}.

In this paper, we show how that assumption can be lifted by switching from
maximum likelihood estimation to \emph{variational inference} (VI), i.e., we
approximate the posterior distribution of the model by optimising the parameters
of a simpler distribution to make it similar to the posterior. This approach can
prove useful in three major ways:
\begin{enumerate}
\item Variance estimates can be used to guide what data should be collected
  next, i.e., if the rewards of some states have abnormally high variance, we
  might want to expose the model to more data visiting those and surrounding
  states.
\item Variances estimates can also be used to judge whether we can trust the
  predictions of the model or, perhaps, the model could benefit from some
  adjustments or more data.
\item By adopting a more Bayesian approach we can set prior distributions on
  random variables. This can help prevent overfitting as well as encode
  additional information about the reward function
  \cite{DBLP:conf/uai/JinDAS17}.
\end{enumerate}

% TODO: could raise intractable posterior as the main reason, but would have to
% rephrase the whole section
No fully Bayesian nonlinear IRL model has been proposed yet. We will uncover one
reason for this when we set up our IRL model for VI in Section~\ref{sec:model}.
Applying VI to this model in a theoretically sound way (without introducing
unjustified simplifying assumptions) requires us to use Lebesgue's dominated
convergence theorem in order to derive the gradient of the expectation of the
MDP value function---we provide a complete proof for this (and some other
results) in Section~\ref{sec:proof}. Finally, in Section~\ref{sec:evaluation} we
demonstrate the model's correctness and properties experimentally using several
example scenarios.

@@

\section{Background} % 0.5-1 page
% TODO: edit this: we can probably extend this (and the introduction)

We argue that developing an IRL model with a variational approximation of the
full posterior distribution has eluded previous research in the area for two
reasons. First, it is not directly relevant to the goal of producing accurate
point estimates of rewards. However, we described several key benefits in the
previous section, and we postulate how this work could affect the wider fields
of reinforcement learning and artificial intelligence in
Section~\ref{sec:conclusions}. Second, as it will become apparent in
Sections~\ref{sec:model} and \ref{sec:proof}, optimising such a VI model
involves estimates of gradients, establishing correctness of which requires a
significant amount of work.

VI has come a long way from the initial \emph{mean field} approach where each
variable is approximated by a univariate Gaussian, independent of all other
variables. Recent years have brought improvements in both computational
complexity and flexibility of the variational approximation
\cite{blei2017variational}. In this section, we will mention some of the most
important and relevant developments in the field, while referring the interested
reader to recent survey articles
\cite{blei2017variational,DBLP:journals/corr/abs-1711-05597} for a broader
overview of the area.

A major development in approximating complex distributions came in the form of
\emph{normalizing flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. A sequence of such
functions can transform a simple initial distribution (e.g., uniform or
Gaussian) into an arbitrarily complex one.
%A major challenge in applying this
%work to IRL is related to variational parameters being used to compute the MDP
%value function, i.e., how can we take the derivative of the value function with
%respect to such a variational parameter? Alternatively, perhaps one can
%construct a different model that would make this question moot.

While these and many other advancements in VI promise great improvements, there
are hidden challenges when it comes to applying them to VI. Many suggested
improvements implicitly impose restrictions on the model that are unlikely to be
satisfied in the context of IRL, e.g., the evidence of the model being a
Gaussian. Nevertheless, the work outlined here seems adaptable to the IRL
setting and can make variational approximations both more accurate and faster.

\section{Problem Description} \label{sec:problem}

@@

\begin{definition}[IRL]
  Given an MDP without rewards $\mathcal{M} \setminus \{ \mathbf{r} \}$, an
  $|\mathcal{S}| \times d$ feature matrix $\mathbf{X}$ (where $d$ is the number
  of features), and a set of \emph{demonstrations} $\mathcal{D} = \{\zeta_i
  \}_{i=1}^N$, where each demonstration $\zeta_i = \{ (s_{i,t}, a_{i,t})
  \}_{t=1}^T$ is a multiset of state-action pairs that represents optimal (or
  near-optimal) actions, find the reward function that maximises the probability
  of observing the demonstrations, i.e.,
  \[
    \argmax_{\mathbf{r}} p(\mathcal{D} \mid \mathbf{r}).
  \]
\end{definition}

\begin{example}
  Figure~\ref{fig:example} shows a possible setup for IRL. In this case,
  $\mathcal{S} = \{ s_1, s_2, s_3, s_4 \}$, $\mathcal{A} = \{ l, r, u, d \}$,
  \begin{align*}
    \mathcal{T}(s, a, s') &= \begin{cases}
      1 & \text{if } (s, a, s') = (s_1, r, s_2) \\
      1 & \text{if } (s, a, s') = (s_1, d, s_4) \\
      1 & \text{if } (s, a, s') = (s_2, l, s_1) \\
      1 & \text{if } (s, a, s') = (s_2, d, s_3) \\
      1 & \text{if } (s, a, s') = (s_3, l, s_4) \\
      1 & \text{if } (s, a, s') = (s_3, u, s_2) \\
      1 & \text{if } (s, a, s') = (s_4, r, s_3) \\
      1 & \text{if } (s, a, s') = (s_4, u, s_1) \\
      0 & \text{otherwise,}
    \end{cases} \\
      \mathbf{X} &= \begin{blockarray}{ccc}
        & f_1 & f_2 \\
        \begin{block}{c[cc]}
          s_1 & 1 & 2 \\
          s_2 & 2 & 2 \\
          s_3 & 2 & 1 \\
          s_4 & 1 & 1 \\
        \end{block}
      \end{blockarray},
  \end{align*}
  and $\gamma \in [0, 1)$ can be assigned any value. The distribution of example
  state-action pairs into separate demonstrations is irrelevant to our model
  (and possible all other models), so we can keep each pair in its own set as
  follows:
  \[
    \mathcal{D} = \{ \{(s_1, d)\}, \{(s_2, l)\}, \{(s_4, u)\}, \{(s_4, u)\} \}.
  \]

@@

A common way to scale GPs to larger data sets is by selecting $m$ points in the
feature space--called \emph{inducing points}---and focus most of the training
effort on them \cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be
the $m \times d$ matrix of features at inducing points, and let $\mathbf{u}$ be
the rewards at those states. Then the full joint probability distribution can be
factorised as
\begin{equation} \label{eq:full}
  \pfull = p(\mathbf{u}) \times p(\mathbf{r} \mid \mathbf{u}) \times
  p(\mathcal{D} \mid \mathbf{r}),
\end{equation}
where $p(\mathbf{u}) = \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu)$ is the GP
prior \cite{DBLP:books/lib/RasmussenW06}. The GP posterior is then a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11} defined as
\begin{equation} \label{eq:r}
  p(\mathbf{r} \mid \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru).
\end{equation}
The matrices such as
$\Kru$ are called \emph{covariance matrices} (also \emph{kernel matrices} and
\emph{Gram matrices}) and are defined as
$[\Kru]_{i,j} = k(\mathbf{x}_{\mathbf{r},i},
\mathbf{x}_{\mathbf{u},j})$, where $\mathbf{x}_{\mathbf{r},i}$ and
$\mathbf{x}_{\mathbf{u},j}$ denote feature vectors for the $i$th state in
$\mathcal{S}$ and the $j$th inducing point, respectively
\cite{DBLP:conf/uai/JinDAS17}.

We can then use VI in order to approximate
$p(\mathbf{u},~\mathbf{r}~\mid~\mathcal{D})$. This is done by minimising the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation. KL divergence (asymmetrically) measures the
difference between two probability distributions. Let $\approximation$ denote
our approximation. The KL divergence is then defined as
\cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}_{\approximation}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}_{\approximation}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}_{\approximation}[\log p(\mathcal{D})].
\end{align*}
The last term is both hard to compute and constant with respect to
$\approximation$, so we can remove it from our optimisation objective. The
negation of what remains is known as the \emph{evidence lower bound} (ELBO) and
is defined as \cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}_{\approximation}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}
Thus, instead of minimising KL divergence, we focus on maximising $\mathcal{L}$
by optimising the values of parameters to be defined in Section~\ref{sec:model}.
Note that hereafter we drop the subscript notation, as all expected values will
be with respect to $\approximation$.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[latent] (u) {$\mathbf{u}$};
    \node[latent, right=of u] (r) {$\mathbf{r}$};
    \node[obs, right=of r] (D) {$\mathcal{D}$};
    \node[const, below=of r] (lambda) {$\bm\lambda$};
    \node[const, below=of u] (mu) {$\bm\mu$};
    \node[const, left=of mu] (B) {$\mathbf{B}$};
    \edge {mu, B} {u};
    \edge {lambda, u} {r};
    \edge {r} {D};
  \end{tikzpicture}
  \caption{Our VI problem expressed as a (simplified)
    Bayesian network. The only observed variable (representing the
    demonstrations) is in a grey circle, modelled latent variables are in white
    circles, and the variational parameters are at the bottom.}
  \label{fig:graphical_model}
\end{figure}

@@

Firstly, let us return to Equation~\eqref{eq:elbo} and write
\[
  \mathcal{L} = \mathbb{E}[\log\pfull] - \mathbb{E}[\log\approximation].
\]
By substituting in Equations~\eqref{eq:full} and \eqref{eq:approximation}, we get
\begin{align*}
  \mathcal{L} &= \mathbb{E}[\log p(\mathbf{u}) + \log p(\mathbf{r} \mid \mathbf{u}) + \log p(\mathcal{D} \mid \mathbf{r})] \\
              &- \mathbb{E}[\log q(\mathbf{u}) + \log q(\mathbf{r} \mid \mathbf{u})].
\end{align*}
Since $q(\mathbf{r} \mid \mathbf{u}) = p(\mathbf{r} \mid \mathbf{u})$, they
cancel each other out. Also notice that
\begin{align*}
  \mathbb{E}[\log p(\mathbf{u}) - \log q(\mathbf{u})] &= -\DKL{q(\mathbf{u})}{p(\mathbf{u})} \\
                                                      &= -\frac{1}{2} (\tr (\Kuu^{-1}\bm\Sigma) + \bm\mu^\intercal\Kuu^{-1}\bm\mu - m \\
                                                      &+ \log |\Kuu| - \log |\bm\Sigma|),
\end{align*}
by the definition of KL divergence between two multivariate Gaussians
\cite{kl}. Hence,
\begin{align*}
  \mathcal{L} &= \mathbb{E}\left[ \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right).
\end{align*}
Using the expressions for $Q_{\mathbf{r}}$ we get
\begin{align*}
  \mathcal{L} &= \mathbb{E}\left[\sum_{i=1}^N \sum_{t=1}^T r(s_{i,t}) - \V(s_{i,t}) + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t}, s')\V(s') \right] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right).
\end{align*}
We can simplify $\sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t})$ by defining a new vector
$\mathbf{t} = (t_1, \dots, t_{|\mathcal{S}|})^\intercal$, where $t_i$ is the
number of times the state associated with reward $r_i$ has been visited
across all demonstrations. Then,
\begin{align*}
  \mathbb{E} \left[ \sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t}) \right] &= \mathbb{E}[\mathbf{t}^\intercal\mathbf{r}] = \mathbf{t}^\intercal\mathbb{E}[\mathbf{r}] \\
                                                               &= \mathbf{t}^\intercal\mathbb{E}\left[\Kru^\intercal\Kuu^{-1}\mathbf{u}\right] = \mathbf{t}^\intercal\Kru^\intercal\Kuu^{-1}\bm\mu.
\end{align*}
This allows us to simplify $\mathcal{L}$ to
\begin{align*}
  \mathcal{L} &= \mathbf{t}^\intercal\Kru^\intercal\Kuu^{-1}\bm\mu - \mathbb{E}[v] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right),
\end{align*}
where
\[
  v = \sum_{i=1}^N \sum_{t=1}^T \V(s_{i,t}) - \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s_{i,t}, a_{i,t}, s')\V(s').
\]
We will need to optimise this expression in order to train our model.

\section{Theoretical Justification} \label{sec:proof}

The typical way to optimise a quantity (the ELBO, in this case) involves
computing its gradient. Unfortunately, the term $\mathbb{E}[v]$ in $\mathcal{L}$
complicates the situation. Indeed, in order to take the derivative of an
expected value, certain conditions must be satisfied. Without proving the
validity of our algorithm, we risk building it on faulty assumptions and having
it misbehave---perhaps all the time, perhaps only in certain situations. The
goal of this section is to show how Lebesgue's dominated convergence theorem can
be applied to our model in order to derive $\nabla\mathbb{E}[v]$, i.e., the
derivative of $\mathbb{E}[v]$ with respect to our variational parameters
$\bm\mu$, $\mathbf{B}$, and $\bm\lambda$. This technique is inspired by black
box VI \cite{DBLP:conf/aistats/RanganathGB14}, but takes a more detailed look at
the problem and requires significantly more work to prove correctness. After
showing that the theorem applies to our situation, we can then estimate
$\nabla\mathbb{E}[v]$ with
\begin{align*}
  \nabla\mathbb{E}[v] &= \nabla \iint v q(\mathbf{r} \mid \mathbf{u}) q(\mathbf{u})\dx \\
                      &= \iint \nabla[v q(\mathbf{r} \mid \mathbf{u}) q(\mathbf{u})]\dx \\
                      &= \iint \frac{\nabla[v q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})]}{q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})} q(\mathbf{r} \mid \mathbf{u}) q(\mathbf{u})\dx \\
                      &\approx \frac{1}{S} \sum_{s=1}^S \frac{\nabla[v q(\mathbf{r}_s \mid \mathbf{u}_s)q(\mathbf{u}_s)]}{q(\mathbf{r}_s \mid \mathbf{u}_s)q(\mathbf{u}_s)},
\end{align*}
which can be computed by drawing $S$ samples $\{(\mathbf{u}_s,
\mathbf{r}_s)\}_{s=1}^S$ from $\approximation$.

Our main goal is Theorem~\ref{thm:main}, which allows us to move differentiation
inside the integral. In order to prove it, we use a number of intermediate
results. We begin by introducing a few important tools and techniques in
Section~\ref{sec:math_preliminaries}. In Section~\ref{sec:derivatives}, we state
a few derivatives of PDFs and covariance matrices and bound their values with
some easy-to-work-with polynomials. In Section~\ref{sec:propositions}, we
provide a sketch proof of the measurability of the MDP value function as well as
new upper and lower bounds on their values. After another quick lemma, the main
proof of the paper is in Section~\ref{sec:the_proof}. See
Figure~\ref{fig:theorems} for an overview of how these results fit together.

@@

\subsection{Differentiability} \label{sec:the_proof}
